---
title: "Module 1: Tabular Data"
subtitle: "Working with larger-than-RAM data using duckdbfs"
author: "ESPM 288"
format: html
---

## Introduction

In this module, we will explore high-performance workflows for tabular data. We will use `duckdbfs` to work with datasets that are larger than available RAM by leveraging DuckDB's streaming and remote file access capabilities.

## Case Study: Global Supply Chains

We will be working with [EXIOBASE 3.8.1](https://source.coop/youssef-harby/exiobase-3), a global Multi-Regional Input-Output (MRIO) database. This dataset tracks economic transactions between sectors and regions, along with their environmental impacts (emissions, resource use, etc.).

**Data description:**
- **Coverage**: 44 countries + 5 rest-of-world regions.
- **Timeframe**: 1995–2022.
- **Content**: Economic transactions (Z matrix), final demand (Y matrix), and environmental stressors (F matrix).
- **Format**: Cloud-optimized Parquet, partitioned by year and matrix type.

## Setup

```{r}
library(duckdbfs)
library(dplyr)

```

## Exercise 1: connecting to remote data

We can open the entire dataset without downloading it using `open_dataset()`. The data is hosted on Source Cooperative. The `**` pattern allows recursive scanning of the partitioned parquet files.

```{r}
# Remote S3 path to EXIOBASE 3 (Source Cooperative)

duckdbfs::duckdb_secrets(
    key = "",
    secret = "",
    endpoint = "s3.amazonaws.com",
    region = "us-west-2"
)
s3_url <- "s3://us-west-2.opendata.source.coop/youssef-harby/exiobase-3/4588235/parquet/year=*/format=ixi/matrix=F_impacts/**"

# Open the dataset lazily
exio <- open_dataset(s3_url)

# View the schema (column names and types) without reading data
glimpse(exio)
```

## Exercise 2: Efficient Filtering

The dataset is large. We should filter *before* collecting any data into R.

```{r}
exio |>
    filter(year == 2022, region == "US") |>
    head() |> # view the first 6 rows
    collect()
```

> **Task**
: Construct a query to find the top 5 sectors in the US by CO2 emissions in 2022. Remember to check the column names in `exio` to find the appropriate emissions flow.

```{r}
#a query written based on the task above, used the chatbot and the info provide in the glimpse. schema problem was in the initial url

exio |>
  distinct(stressor) |>
  collect()


exio |>
  filter(year == 2022, region == "US") |>
  group_by(sector) |>
  summarize(total_co2 = sum(.data$value, na.rm = TRUE), .groups = "drop") |>
  slice_max(order_by = total_co2, n = 5) |>
  collect()
```

## Visualization: Top 10 Sectors by CO2 Emissions

```{r}
library(ggplot2)

# Get top 10 sectors by CO2 emissions
top10_sectors <- exio |>
  filter(year == 2022, region == "US") |>
  group_by(sector) |>
  summarize(total_co2 = sum(.data$value, na.rm = TRUE), .groups = "drop") |>
  slice_max(order_by = total_co2, n = 10) |>
  collect()

# Create the visualization
p <- ggplot(top10_sectors, aes(x = reorder(sector, total_co2), y = total_co2)) +
  geom_col(fill = "#FF10F0") +
  coord_flip() +
  labs(
    title = "Top 10 Sectors by CO2 Emissions in the US (2022)",
    x = "Sector",
    y = "Total CO2 Emissions"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

# Display the plot
print(p)

# Export the visualization
ggsave("top10_sectors_co2.png", plot = p, width = 10, height = 6, dpi = 300)
```

## Forecasting: Predicting CO2 Emissions by Country (2050-2055) Using MARSS

```{r}

#this has got some problems, predictions not working. Maybe not a good fit for time series data? predicting too far in advance given that the og data is a short time period?
library(tidyr)
library(purrr)
library(MARSS)

# Get historical CO2 emissions by country and year
historical_data <- exio |>
  group_by(region, year) |>
  summarize(total_co2 = sum(.data$value, na.rm = TRUE), .groups = "drop") |>
  collect() |>
  arrange(region, year)

# Check data structure
cat("Number of regions:", length(unique(historical_data$region)), "\n")
cat("Year range:", min(historical_data$year), "to", max(historical_data$year), "\n")

# Function to forecast using MARSS for a single country
forecast_country_marss <- function(region_name, country_data) {
  cat("Processing:", region_name, "\n")
  
  tryCatch({
    # Prepare time series data - needs to be a matrix
    ts_data <- matrix(country_data$total_co2, nrow = 1)
    
    # Simple univariate MARSS model
    model_list <- list(
      B = matrix(1),  # Random walk
      U = matrix(0),  # No drift
      Q = matrix("q"),  # Process variance
      Z = matrix(1),  # Observation model
      A = matrix(0),  # No offset
      R = matrix("r")  # Observation variance
    )
    
    # Fit MARSS model
    fit <- MARSS(ts_data, model = model_list, silent = TRUE)
    
    # Get the last state estimate
    last_state <- fit$states[, ncol(fit$states)]
    
    # Simple projection: use the estimated trend from the model
    # Get parameter estimates
    B_est <- coef(fit, type = "matrix")$B
    
    # Project forward 33 years (2023-2055)
    n_ahead <- 33
    predictions <- numeric(n_ahead)
    current_state <- last_state
    
    for (i in 1:n_ahead) {
      current_state <- B_est * current_state
      predictions[i] <- current_state
    }
    
    # Extract predictions for 2050-2055 (years 28-33)
    predictions_2050_2055 <- predictions[28:33]
    
    cat("  Success! Predictions range:", round(min(predictions_2050_2055)), 
        "to", round(max(predictions_2050_2055)), "\n")
    
    data.frame(
      region = region_name,
      year = 2050:2055,
      predicted_co2 = predictions_2050_2055
    )
  }, error = function(e) {
    cat("  Error:", conditionMessage(e), "\n")
    # Return NULL instead of NA so we can filter it out
    return(NULL)
  })
}

# Apply MARSS forecasting to each country
cat("\n=== Starting MARSS Forecasting ===\n")

predictions_list <- historical_data |>
  group_by(region) |>
  group_split() |>
  map(function(df) {
    region_name <- df$region[1]
    forecast_country_marss(region_name, df)
  })

# Combine all predictions, removing NULL entries
predictions_by_country <- bind_rows(predictions_list[!sapply(predictions_list, is.null)])

cat("\n=== Forecasting Complete ===\n")
cat("Successfully predicted for", length(unique(predictions_by_country$region)), "regions\n\n")

# Calculate average emissions for 2050-2055 by country
avg_predictions <- predictions_by_country |>
  group_by(region) |>
  summarize(avg_co2_2050_2055 = mean(predicted_co2), .groups = "drop") |>
  arrange(desc(avg_co2_2050_2055))

# Display top 10 countries
cat("\n=== Top 10 Countries by Predicted Average CO2 (2050-2055) ===\n")
print(avg_predictions |> head(10))

# Display all predictions
cat("\n=== All Predictions by Year ===\n")
print(predictions_by_country |> arrange(region, year))

# Visualize predicted average emissions
p_forecast <- ggplot(avg_predictions |> head(15), 
                     aes(x = reorder(region, avg_co2_2050_2055), y = avg_co2_2050_2055)) +
  geom_col(fill = "#FF10F0") +
  coord_flip() +
  labs(
    title = "Predicted Average CO2 Emissions by Country (2050-2055)",
    subtitle = "Based on MARSS state-space model from historical data (1995-2022)",
    x = "Country/Region",
    y = "Predicted Average CO2 Emissions"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

# Display the plot
p_forecast

# Export the forecast visualization
ggsave("predicted_co2_2050_2055.png", plot = p_forecast, width = 10, height = 6, dpi = 300)

# Export predictions to CSV
write.csv(avg_predictions, "co2_predictions_2050_2055.csv", row.names = FALSE)
```
